---
title: "Assignment_3B"
author: "Nana Kwasi Danquah"
format: html
editor: visual
---

## 1. Project Approach

My goal is to clean up sales numbers so we can see the real trend. We do this by looking at two different timeframes at the same time:

**Partitioning:** I will look at each item separately. This is crucial so that "Product A" doesn't mess up the averages for "Product B."

**The Year-to-Date (YTD) Average:** I will use an Expanding Window. It starts at the first row and gets bigger as it goes down, calculating a "running average" of everything seen so far.

**The 6-Day Moving Average:** I will use a Sliding Window. It only looks at the "Current Day + the 5 days before it." This is used to smooth out daily and see the current trend.

## 2. Data Transformation (dplyr)

This section performs the actual math. I use the `dplyr` library to handle the "windows" and the `zoo` library to calculate the rolling 6-day average.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(zoo)

# Loading the dataset you provided
sales_df <- read_csv("sales_with_window_functions.csv")

# Calculating the Window Functions
final_data <- sales_df %>%
  group_by(Product) %>%      # Partitioning: Keep A and B separate
  arrange(Date) %>%          # Ordering: Ensure the timeline is correct
  mutate(
    # The Expanding Window (YTD)
    YTD_Average = cummean(Units_Sold),
    
    # The Sliding Window (6-Day)
    # k=6 means a 6-day window; align='right' looks at today + 5 days back
    Six_Day_Moving_Avg = rollmean(Units_Sold, k = 6, fill = NA, align = "right")
  ) %>%
  ungroup()

# Displaying the results table
knitr::kable(final_data, digits = 2)
```

## 3. Visualizing the Results

By plotting the data, I can see exactly what the window functions did. The individual points are the "noisy" daily sales, while the solid lines represent the 6-Day Moving Average.

```{r}
ggplot(final_data, aes(x = as.Date(Date), color = Product)) +
  geom_point(aes(y = Units_Sold), alpha = 0.3) +
  # We updated 'size' to 'linewidth' to fix the warning
  geom_line(aes(y = Six_Day_Moving_Avg), linewidth = 1.2) + 
  labs(
    title = "Sales Trends: 6-Day Moving Average",
    subtitle = "The bold lines smooth out the daily jumps to show the true direction",
    x = "Date",
    y = "Units Sold"
  ) +
  theme_minimal()
```

## 4. Final Conclusion

Using Window Functions allows us to see that even though sales might drop on a specific Monday, the 6-Day Moving Average shows if the business is actually healthy.

**Smartwatch A** shows high volume but a very steady trend line.

**Smartwatch B** is lower in volume but shows a very consistent "Pulse," making it a predictable product for inventory planning.
