---
title: "Analysis of Student Performance"
author: "Nana Kwasi Danquah"
format: html
execute:
  echo: true
---

## Approach

I created a beachhead, where i downloaded a csv file from kaggle dateset about student performance.

[Source: https://www.kaggle.com/datasets/nadeemajeedch/students-performance-10000-clean-data-eda?resource=download](https://www.kaggle.com/datasets/nadeemajeedch/students-performance-10000-clean-data-eda?resource=download)

## Introduction and Planned Approach

In order to determine the relationship between academic achievement and demographic, educational, and behavioral characteristics, this research examines a dataset of 10,000 students. Subject-level scores and an overall total score are used to measure student results, providing both aggregate and specific viewpoints on performance.

The analytical approach is staged on purpose. First, exploratory data analysis (EDA) is used to evaluate the data's quality and structure. Understanding distributions, spotting abnormalities, and analyzing performance differences across important student attributes including gender, parental education, meal status, and test preparation involvement are all done in this step.

The study proceeds to explanatory modeling once a comprehensive grasp of the data has been established. The objective is to measure and analyze the connections between student characteristics and academic results rather than concentrating only on prediction accuracy. To guarantee that results may be effectively addressed in an educational setting, simple, comprehensible models are given priority.

This Quarto file documents every stage of the workflow, from data import to final findings, making it completely replicable.

## Anticipated Data Challenges

This dataset has a number of intrinsic difficulties. First, several variables—especially test scores—had missing values, which could have an impact on modeling outcomes and summary statistics. These missing values must be treated carefully to avoid biased findings.

Second, a number of variables (such as parental education and race/ethnicity) are categorical with different degrees of granularity. Model stability and interpretation may be impacted by choices made regarding the encoding of these variables.

Third, especially for high achievers, student performance metrics may show skewness or ceiling effects. This needs to be checked diagnostically because it can go against modeling assumptions.

Lastly, any links found should be understood as associations rather than causal effects because the data are observational.


```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(scales)

# 1. Load the data 
students <- read_csv("Student_performance_10k.csv")

# 2. Data Cleaning Workflow
# We convert math_score to numeric because it was imported as text.
# We then remove any rows with missing values (NA) to ensure our charts are accurate.
students <- students %>%
  mutate(math_score = as.numeric(as.character(math_score))) %>%
  drop_na()

# 3. Organizing Categories
# We manually order education levels so they appear logically (from low to high) in charts.
students$parental_level_of_education <- factor(students$parental_level_of_education, 
    levels = c("some high school", "high school", "some college", 
               "associate's degree", "bachelor's degree", "master's degree"))
```

## Data Overview
We start by examining the cleaned dataset. After cleaning, we have a consistent set of scores for Math, Reading, Writing, and Science.

```{r}
head(students)
```

## Analysis of Top Achievers

I filtered the data to identify "High Scorers"—students with a total score greater than 300. This count helps us see what percentage of the student body is performing at an elite level.

```{r}
high_scores <- students %>%
  filter(total_score > 300)

# Display the number of students scoring above 300
cat("Number of students with scores over 300:", nrow(high_scores))
```

## Visualizing Performance Drivers

## 1. Impact of Parental Education
The boxplot below shows the distribution of total scores based on the parents' highest level of education.

```{r}
ggplot(students, aes(x = parental_level_of_education, y = total_score, fill = parental_level_of_education)) +
  geom_boxplot(outlier.shape = 21, outlier.alpha = 0.3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Total Score vs. Parental Education",
       subtitle = "Note: Scores tend to increase with higher parental education levels.",
       x = "Parental Education Level",
       y = "Total Score") +
  guides(fill = "none")
```

## 2. Effectiveness of Test Preparation

We compare the average total scores of students who completed a preparation course versus those who did not.

```{r}
prep_summary <- students %>%
  group_by(test_preparation_course) %>%
  summarise(mean_total = mean(total_score))

ggplot(prep_summary, aes(x = factor(test_preparation_course), y = mean_total, fill = factor(test_preparation_course))) +
  geom_bar(stat = "identity", width = 0.5) +
  theme_minimal() +
  scale_x_discrete(labels = c("0" = "None", "1" = "Completed")) +
  labs(title = "Average Score by Test Prep Course",
       x = "Course Completion",
       y = "Average Total Score") +
  guides(fill = "none")
```

## Correlation Between Subjects

To see if students who are good at one subject are likely to be good at others, we create a correlation heatmap

```{r}
# Step 1: Create a correlation matrix
cor_data <- students %>%
  select(math_score, reading_score, writing_score, science_score) %>%
  cor()

# Step 2: Convert to a long format for ggplot (modern way to do it without reshape2)
cor_long <- as.data.frame(cor_data) %>%
  rownames_to_column(var = "Subject1") %>%
  pivot_longer(cols = -Subject1, names_to = "Subject2", values_to = "Correlation")

# Step 3: Plot the heatmap
ggplot(cor_long, aes(x = Subject1, y = Subject2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2))) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5) +
  theme_minimal() +
  labs(title = "Subject Score Correlation Heatmap",
       subtitle = "1.00 represents a perfect positive correlation.")
```

## Final Insights
## 1.Prep Courses Work: 
On average, students who completed the test preparation course scored higher than those who did not.

## 2.Parental Education Influence: 
There is a clear positive trend; students with parents holding Master's or Bachelor's degrees achieve higher median scores.

## 3.The Literacy Connection: 
Reading and Writing scores have the highest correlation (~0.95), suggesting that these two skills are nearly identical in their performance requirements.

## 5.Workflow Improvement: 
By handling the math_score as a numeric type and filtering out NAs at the start, we eliminated the warnings and errors that were hindering the analysis.