---
title: "Assignment_3A"
author: "Nana Kwasi Danquah"
format: html
editor: visual
---

## 1. Project Approach

This project's goal is to use survey data to construct a **Global Baseline Estimate (GBE)** recommendation system. It is a statistical baseline that produces high-quality recommendations by correcting for individual and item-level biases, in contrast to elaborate machine learning models that require significant processing power.

### Approach

My approach follows a 5-step data science pipeline:

1.  **Normalization and Ingestion:** The "Wide" matrix is imported from the spreadsheet. I pivot the data into a "Long" format (Tidy Data) because the technique necessitates mathematical operations across rows and columns.
2.  **Sanitization of Data:** To guarantee calculation accuracy, I detect and manage non-numeric noise (such as the "?" entries present in the raw survey).
3.  **Global Parameterization (**$\mu$): For each forecast, I compute the "Global Mean" as the initial value.
4.  **Bias Decomposition:**
    -   **User Bias (**$b_u$): To distinguish between "harsh" and "generous" critics, I calculate the deviation of a critic's average rating from the worldwide average.
    -   **Movie Bias (**$b_i$): I quantify the ratio of "blockbusters" to "flops" based on how much a film's average rating differs from the global average.
5.  **Predictive Synthesis:** To fill all gaps in the survey matrix, I use the formula $b_{u,i} = \mu + b_u + b_i$.

## 2. Ingestion and Normalization

I import the "Wide" matrix directly from the Excel file. I use the read_excel function, which correctly parses the binary format. I then pivot the data into a "Long" format (Tidy Data) by keeping the first column (the Critics) as an identifier and collapsing the movie columns.

```{r ingestion, include=FALSE}
if (!require("readxl")) install.packages("readxl")
library(readxl)
library(tidyverse)
library(knitr)

# Load the Excel file
file_path <- "MovieRatings.xlsx" 

raw_data <- read_excel(file_path)

# Standardize the first column name to 'Critic'
colnames(raw_data)[1] <- "Critic"

# Pivot to Long format (Tidy Data)
# 'cols = -1' tells R to pivot everything EXCEPT the first column
tidy_ratings <- raw_data %>%
  pivot_longer(
    cols = -1, 
    names_to = "Movie", 
    values_to = "Raw_Value"
  )

kable(head(tidy_ratings), caption = "Normalized Long Format Data")
```

## 3. Data Sanitization

I identify and handle non-numeric noise (e.g., the "?" entries found in the survey). In Excel, these often import as characters. I convert these to numeric values; symbols like "?" or text strings will automatically become NA. I then filter out these NA rows to create a clean set of numbers for our calculation.

```{r}
clean_ratings <- tidy_ratings %>%
  # Force conversion to numeric (turns '?' and blanks into NA)
  mutate(Rating = suppressWarnings(as.numeric(as.character(Raw_Value)))) %>%
  # Remove rows that do not have a valid numeric rating
  filter(!is.na(Rating)) %>%
  select(Critic, Movie, Rating)

kable(head(clean_ratings), caption = "Sanitized Numeric Ratings")
```

## 4. Global Parameterization ($\mu$)

I calculate the Global Mean ($\mu$). This is the single most important starting point: the average of every valid rating in the entire dataset.

```{r}
mu <- mean(clean_ratings$Rating)
cat("The Global Parameter (mu) is:", round(mu, 4))
```

## 5. Bias Decomposition

I calculate the deviations for critics and movies from the global average: **User Bias (**$b_u$): How much a critic's average rating deviates from $\mu$. Positive values mean they are "generous," negative values mean they are "harsh." **Movie Bias (**$b_i$): How much a movie's average rating deviates from $\mu$. This identifies the "blockbusters" vs. the "flops."

```{r}
# Calculate User Bias (bu)
user_bias <- clean_ratings %>%
  group_by(Critic) %>%
  summarize(b_u = mean(Rating) - mu)

# Calculate Movie Bias (bi)
movie_bias <- clean_ratings %>%
  group_by(Movie) %>%
  summarize(b_i = mean(Rating) - mu)

kable(user_bias, caption = "User Biases (bu)")
kable(movie_bias, caption = "Movie Biases (bi)")
```

## 6. Predictive Synthesis

I apply the formula $b_{u,i} = \mu + b_u + b_i$ to fill every gap in the matrix. This combines the overall average, the specific critic's tendency, and the movie's general appeal to create a customized prediction for missing ratings.

```{r}
# 1. Create a full grid of every Critic and every Movie
full_grid <- expand.grid(
  Critic = unique(raw_data$Critic),
  Movie = unique(clean_ratings$Movie)
)

# 2. Join the biases and calculate the recommendation
final_recommendations <- full_grid %>%
  left_join(user_bias, by = "Critic") %>%
  left_join(movie_bias, by = "Movie") %>%
  mutate(
    Estimate = mu + b_u + b_i,
    # Standardize scale: ensure results stay between 1.0 and 5.0
    Estimate = pmax(1, pmin(5, Estimate))
  )

# 3. Pivot back to the original Wide Matrix format for the report
recommendation_matrix <- final_recommendations %>%
  select(Critic, Movie, Estimate) %>%
  pivot_wider(names_from = Movie, values_from = Estimate)

kable(recommendation_matrix, digits = 2, caption = "Completed Global Baseline Recommendation Matrix")
```
