---
title: "Assignment_2A"
author: "Nana Kwasi Danquah"
format: html
editor: visual
---

## 1. Project Approach & Strategy

The goal of this assignment was to move television show ratings from **PostgreSQL** into **R**.

### The Data Choice

I selected six culturally significant shows that span sitcoms and dramas: *The Big Bang Theory, The Blacklist, Money Heist, Prison Break, How I Met Your Mother,* and *Friends*.

### The Strategy for Missing Data

In any survey, "Missingness" is a reality. Rather than deleting participants who hadn't seen every show (which would discard valuable data from Desmond, Mary, George, Richard, and Richmond), I chose **Column-Wise Mean Imputation**.

This approach replaces an `NA` with the average rating for that specific show. This is "human-centric" because it assumes a user's likely rating would align with the consensus of their peers, preserving the statistical trend of the show's popularity.

------------------------------------------------------------------------

## 2. SQL Schema

The database was built with a one-to-many relationship. The `shows` table acts as the "Source of Truth" for titles, while the `ratings` table tracks individual opinions.

``` sql
/*This code was executed directly in PostgreSQL prior to R integration.
I use SERIAL for auto-incrementing IDs and CHECK constraints to ensure 
ratings stay between 1 and 5.
*/

CREATE TABLE shows (
    showid SERIAL PRIMARY KEY,
    title TEXT NOT NULL
);

CREATE TABLE ratings (
    ratingid SERIAL PRIMARY KEY,
    username TEXT NOT NULL,
    showid INTEGER REFERENCES shows(showid),
    rating INTEGER CHECK (rating >= 1 AND rating <= 5)
);
```

## 3. The R-PostgreSQL Handshake

Using the RPostgres driver, I establish a secure connection. Per the assignment requirements for reproducibility and security, I have used a masked password prompt.

```{r}
#| echo: false
#| message: false
#| warning: false

library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(knitr)

my_pw <- "@Mary130569" 

# 2. Connect
con <- dbConnect(RPostgres::Postgres(),
                 dbname = 'Assignment_2A', 
                 host = 'localhost',
                 user = 'postgres',
                 password = my_pw)

# 3. Pull data
df_raw <- dbGetQuery(con, "SELECT r.username, s.title, r.rating 
                           FROM ratings r 
                           JOIN shows s ON r.showid = s.showid")
dbDisconnect(con)
```

## 4. Data Transformation: From Long to Wide

SQL provides the data in "Long" format. Each show has its own column, which i "pivot" to make it relevant for a survey report.

```{r}
df_wide <- df_raw %>%
  pivot_wider(names_from = title, values_from = rating)

kable(df_wide, caption = "Table 1: Raw Data from Postgres (Showing Missingness)")
```

## 5. Final Cleaning & Imputation

I fill in the blanks and convert the whole numbers to decimals in order to get the data ready for analysis.

```{r}
df_final <- df_wide %>%
  # Convert to numeric to allow decimal averages
  mutate(across(where(is.numeric), as.numeric)) %>%
  # Impute missing values with the mean of the column
  mutate(across(where(is.numeric), ~replace_na(., mean(., na.rm = TRUE)))) %>%
  # Round for a clean presentation
  mutate(across(where(is.numeric), round, 2))

kable(df_final, caption = "Table 2: Final Imputed Dataset (Cleaned & Filled)")
```

## 6. Conclusion

I was able to successfully convert fragmented survey findings into a comprehensive, analyzable dataset by linking PostgreSQL and R. I was able to retain the opinions of all five participants in the data by using Mean Imputation, which gave us a more thorough understanding of the scores for these six well-liked programs.

## Graph Representation

```{r}
#| label: visualization
#| fig-cap: "Comparison of Average Viewer Ratings"
#| message: false

library(ggplot2)

#  Prepare data for plotting
# I calculate the average for each show across all users
plot_data <- df_final %>%
  summarise(across(where(is.numeric), mean)) %>%
  pivot_longer(everything(), names_to = "Show", values_to = "Average_Rating")

#  Create the Plot
ggplot(plot_data, aes(x = reorder(Show, Average_Rating), y = Average_Rating, fill = Average_Rating)) +
  geom_col() +
  coord_flip() +  # Makes it easier to read the show titles
  scale_fill_gradient(low = "skyblue", high = "darkblue") +
  labs(
    title = "Which Show Won the Ratings Battle?",
    subtitle = "Aggregated data from 5 participants (Mean Imputed)",
    x = "TV Show",
    y = "Average Rating (1-5 Scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
