---
title: "Assignment_2B"
subtitle: "Evaluating Model Performance Across Decision Thresholds"
author: "Nana Kwasi Danquah"
format: html
editor: visual
---

## 1. Introduction

This analysis evaluates a classification model's ability to predict the sex of penguins. We examine the trade-offs between precision and recall at various probability thresholds and compare the results against a baseline null model.

## 2. Approach (Methodology)

Our approach for this project followed a structured data science workflow:

1.  **Baseline Establishment:** I calculated the Null Error Rate to define the minimum performance required for the model to be considered useful.
2.  **Probability Transformation:** I used the raw probability column (`.pred_female`) to generate binary classifications at three specific thresholds (0.2, 0.5, and 0.8).
3.  **Error Classification:** For each threshold, I mapped results into Confusion Matrices to quantify True Positives, False Positives, True Negatives, and False Negatives.
4.  **Multi-Metric Evaluation:** I calculated Accuracy, Precision, Recall, and F1-score to provide a multi-dimensional view of performance beyond simple accuracy.
5.  **Strategic Contextualization:** I interpreted the statistical results into real-world conservation and research scenarios to justify threshold selection.


## 3. Exploratory Analysis & Baseline

```{r}
#| label: setup
#| message: false
#| warning: false
library(ggplot2)
library(dplyr)
library(tidyr)

# Load the dataset
df <- read.csv('penguin_predictions.csv')

# Calculate Null Error Rate
counts <- table(df$sex)
null_accuracy <- max(counts) / nrow(df)
null_error_rate <- 1 - null_accuracy
```


# Actual Class Distribution

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of Actual Sex labels in the dataset."
ggplot(df, aes(x = sex, fill = sex)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Distribution of Actual Class (Sex)",
       y = "Count", x = "Sex") +
  theme_minimal()
```

The Null Error Rate is r round(null_error_rate, 4). This baseline confirms that simply guessing the majority class ("male") would lead to an error 41.94% of the time.


## 4. Performance at Multiple Thresholds

```{r}
#| label: calculate-metrics
thresholds <- c(0.2, 0.5, 0.8)

calc_metrics <- function(t) {
  # Predictions based on threshold
  preds <- ifelse(df$.pred_female >= t, "female", "male")
  
  # Confusion Matrix elements
  tp <- sum(df$sex == "female" & preds == "female")
  fp <- sum(df$sex == "male" & preds == "female")
  tn <- sum(df$sex == "male" & preds == "male")
  fn <- sum(df$sex == "female" & preds == "male")
  
  # Metrics
  accuracy <- (tp + tn) / nrow(df)
  precision <- tp / (tp + fp)
  recall <- tp / (tp + fn)
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  return(data.frame(Threshold = t, TP = tp, FP = fp, TN = tn, FN = fn, 
                    Accuracy = accuracy, Precision = precision, 
                    Recall = recall, F1 = f1))
}

results <- do.call(rbind, lapply(thresholds, calc_metrics))
```

#Performance Metrics Summary

```{r}
#| label: tbl-summary
knitr::kable(results, digits = 4, caption = "Model Performance by Threshold")
```


## 5. Discussion: Strategic Threshold Selection

# The 0.2 Threshold (Prioritization Recall)
When it is expensive to miss a female, a 0.2 threshold is ideal.

Situation: In order to give pre-nesting nutrients, a veterinarian must identify every female penguin. Even if a few males are examined needlessly, using a 0.2 threshold (Recall = 0.9487) guarantees that nearly all females are contacted.

# The 0.8 Threshold (Prioritizing Precision)
When identifying a male as female is expensive, a threshold of 0.8 is acceptable.

Situation: An extremely costly genetic test that is only effective on females is being used by a researcher. They utilize a 0.8 threshold (Precision = 0.9474) to be sure that the birds selected for testing are almost certainly female in order to prevent wasting the expensive kit.