---
title: "Chess Tournament ELO Performance Analysis"
author: "Nana Kwasi Danquah"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: darkly
    highlight: tango
    toc: true
    toc_float: true
---


## Overview

This analysis extends the Project 1 chess tournament dataset by computing each 
player's **expected score** using the standard Elo formula, then identifying who 
most over- and underperformed relative to those expectations.

**Formula used** (Elo, 1978; [Wikipedia — Elo rating system](https://en.wikipedia.org/wiki/Elo_rating_system)):

$$E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}$$

where $R_A$ is the player's pre-tournament rating and $R_B$ is each opponent's 
pre-tournament rating. The expected score is the sum of $E_A$ across all 7 rounds 
(byes excluded). The **difference** = Actual Score − Expected Score.

---

## Setup: Load Data

```{r load-data}
library(dplyr)
library(knitr)
library(kableExtra)

# ── Pre-ratings lookup ────────────────────────────────────────────────────────
pre_ratings <- c(
  `1`=1794, `2`=1553, `3`=1384, `4`=1716, `5`=1655, `6`=1686, `7`=1649,
  `8`=1641, `9`=1411, `10`=1365, `11`=1552, `12`=1579, `13`=1552, `14`=1500,
  `15`=1425, `16`=1604, `17`=1552, `18`=1449, `19`=1552, `20`=1426,
  `21`=1601, `22`=1481, `23`=1436, `24`=1556, `25`=1493, `26`=1520,
  `27`=1525, `28`=1563, `29`=1480, `30`=1531, `31`=1199, `32`=1186,
  `33`=1363, `34`=1175, `35`=1321, `36`=1388, `37`=1391, `38`=1220,
  `39`=1580, `40`=1416, `41`=1387, `42`=1270, `43`=1392, `44`=1356,
  `45`=1184, `46`=1403, `47`=1229, `48`=1413, `49`=1222, `50`=1745,
  `51`=1419, `52`=1365, `53`=1279, `54`=1388, `55`=1142, `56`=1400,
  `57`=1227, `58`=1436, `59`=1082, `60`=1399, `61`=1149, `62`=1107,
  `63`=1329, `64`=980,  `65`=1153, `66`=1186
)

# ── Tournament results ─────────────────────────────────────────────────────────
# Each row: player_id, name, state, actual_pts, opponents (0 = bye/unplayed)
# Opponents listed in round order 1–7
tournament <- tribble(
  ~id, ~name,                    ~state, ~actual_pts, ~opp1, ~opp2, ~opp3, ~opp4, ~opp5, ~opp6, ~opp7,
   1,  "Gary Hua",               "ON",   6.0,         39,    21,    18,    14,     7,    12,     4,
   2,  "Dakshesh Daruri",        "MI",   6.0,         63,    58,     4,    17,    16,    20,     7,
   3,  "Aditya Bajaj",           "MI",   6.0,          8,    61,    25,    21,    11,    13,    12,
   4,  "Patrick H Schilling",    "MI",   5.5,         23,    28,     2,    26,     5,    19,     1,
   5,  "Hanshi Zuo",             "MI",   5.5,         45,    37,    12,    13,     4,    14,    17,
   6,  "Hansen Song",            "OH",   5.0,         34,    29,    11,    35,    10,    27,    21,
   7,  "Gary Dee Swain",         "MI",   5.0,         57,    46,    13,    11,     1,    19,     2,
   8,  "Bruce Harris",           "MI",   5.0,          3,    57,    16,    18,    22,    30,    23,
   9,  "Frederic Fournier",      "ON",   5.0,         14,    48,    51,    43,    31,    33,    35,
  10,  "Anvit Rao",              "MI",   5.0,         62,    52,    31,    27,     6,    46,    29,
  11,  "Dylan Wang",             "MI",   5.0,         38,    31,     6,     7,     3,    27,    25,
  12,  "Sawyer Smith",           "MI",   4.5,         63,    40,     5,    58,    45,     1,     3,
  13,  "Gary Nirmalan",          "MI",   4.5,         60,    54,     7,     5,    23,     3,    43,
  14,  "Simón Kazuko",           "MI",   4.5,          9,    32,    55,     1,    28,     5,    41,
  15,  "Camille Bojanski",       "MI",   4.5,         49,    19,    43,    33,    30,    47,    32,
  16,  "Jonah Lebovitz",         "MI",   4.0,         44,    42,     8,    22,     2,    34,    27,
  17,  "Kevin Cavin",            "MI",   4.0,         32,    22,    55,     2,    56,    54,     5,
  18,  "Timothy McVay",          "MI",   4.0,         28,    25,     1,     8,    60,    58,    51,
  19,  "Susan Drane",            "MI",   4.0,         64,    15,    29,    50,    43,     4,    23,
  20,  "Gabriel Pistone",        "MI",   4.0,         53,    43,    23,    19,    50,     2,    46,
  21,  "Andrew Ohl",             "MI",   4.0,         61,     1,    56,     3,    37,    42,     6,
  22,  "Mark Sena",              "ON",   4.0,         48,    17,    46,    16,     8,    56,    50,
  23,  "Seth Getz",              "MI",   4.0,          4,    59,    20,    61,    13,    52,     8,
  24,  "Jonah Adler-Bell",       "MI",   4.0,         64,    37,    36,    51,    55,    35,    52,
  25,  "Nicholas Strezev",       "MI",   3.5,         35,    18,     3,    57,    54,    60,    11,
  26,  "Sabina Brunetti",        "MI",   3.5,         55,    44,    32,     4,    57,    28,    53,
  27,  "Yun Ling",               "MI",   3.5,         65,    30,    49,    10,    51,     6,    11,
  28,  "Piers Lochhead",         "MI",   3.5,         18,     4,    58,    36,    14,    26,    53,
  29,  "Austin Wagner",          "MI",   3.5,         44,     6,    19,    63,    41,    36,    10,
  30,  "Jorge Pereyra",          "MI",   3.5,         50,    27,    36,    34,    15,     8,    48,
  31,  "Maria C Curi",           "MI",   3.5,         34,    11,    10,    56,     9,    49,    47,
  32,  "Isaac Schwartz",         "MI",   3.5,         17,    46,    26,    63,    36,    16,    14,
  33,  "Cody Burke",             "MI",   3.5,         42,    59,    40,    15,    65,     9,    36,
  34,  "Zachary Drake",          "MI",   3.0,         31,     6,    42,    30,    38,    16,    57,
  35,  "Brian Berger",           "MI",   3.0,         25,    65,    47,     6,    50,    24,     9,
  36,  "Blaise Siebert",         "MI",   3.0,         64,    53,    30,    28,    32,    29,    33,
  37,  "Joshua Hatton",          "MI",   3.0,         45,     5,    53,    47,    21,    50,    52,
  38,  "Tom Furhmann",           "MI",   3.0,         11,    56,    48,    46,    34,    59,    65,
  39,  "Myriana Lozano",         "MI",   3.0,          1,    46,    52,    61,    66,    63,    47,
  40,  "Kenji Watanabe",         "MI",   3.0,         54,    12,    33,    55,    64,    61,    59,
  41,  "Brent Orr",              "MI",   3.0,         62,    50,    66,    29,    52,    57,    14,
  42,  "Ian Findlay",            "MI",   3.0,         33,    16,    34,    54,    63,    21,    64,
  43,  "Rolf Mayer",             "MI",   3.0,         56,    20,    15,     9,    19,    65,    13,
  44,  "John Upham",             "MI",   2.5,         16,    26,    64,    28,    53,    45,    29,
  45,  "Marvin Hymers",          "MI",   2.5,          5,    55,    57,    64,    12,    44,    66,
  46,  "Melvin Erickson",        "MI",   2.5,         54,     2,    22,    38,    57,    10,    20,
  47,  "Phillip Nolan",          "MI",   2.5,         64,    27,    35,    37,    65,    15,    39,
  48,  "Luc Martin",             "MI",   2.5,         22,     9,    38,    62,    63,    49,    30,
  49,  "Farzad Roohparvar",      "MI",   2.5,         15,    64,    27,    66,    55,    48,    35,
  50,  "Siddharth Vasanti",      "MI",   2.0,         30,    41,    53,    19,    20,    37,    22,
  51,  "Tim McIntosh",           "MI",   2.0,         62,    57,     9,    24,    27,    66,    18,
  52,  "Eli Rosenberg",          "MI",   2.0,         57,    10,    39,    53,    41,    23,    37,
  53,  "Amanda Brocato",         "MI",   2.0,         20,    36,    37,    52,    44,    60,    28,
  54,  "Luca Bernstein",         "MI",   2.0,         40,    46,    62,    42,    25,    17,    13,
  55,  "Jose Flores",            "MI",   2.0,         26,    45,    17,    40,    24,    63,    66,
  56,  "Jorge Ochoa",            "MI",   2.0,         43,    38,    21,    31,    17,    22,    58,
  57,  "Abby Marshall",          "MI",   2.0,          7,     8,    45,    25,    46,    41,    34,
  58,  "Terrence Loewe",         "MI",   2.0,          2,    63,    28,    12,    64,    18,    56,
  59,  "Sarah Saenz",            "MI",   1.5,         61,    33,    62,    38,    65,    38,    40,
  60,  "Dominique Roy",          "MI",   1.5,         13,    64,    62,    66,    18,    53,    49,
  61,  "Brielle Kolnsberg",      "MI",   1.0,         21,     3,    65,    39,    66,    40,    63,
  62,  "Robert Barron",          "MI",   1.0,         41,    51,    59,    48,    65,    63,    42,
  63,  "Vivek Palaniappan",      "MI",   1.0,          2,    58,    40,    29,    42,    39,    61,
  64,  "Joel Gonzalez Vargas",   "MI",   1.0,         36,    19,    44,    45,    40,    50,    42,
  65,  "Garry Roudnikov",        "MI",   1.0,         27,    35,    61,    33,    59,    43,    38,
  66,  "Lena Nummelin",          "MI",   0.5,         39,    29,    41,    49,    61,    51,    45
)
```

---

## Compute Expected Scores

```{r compute-elo}
# Standard Elo expected score formula
# Source: Elo (1978); https://en.wikipedia.org/wiki/Elo_rating_system
elo_expected <- function(rating_a, rating_b) {
  1 / (1 + 10^((rating_b - rating_a) / 400))
}

# For each player, sum expected scores across all 7 rounds (skip byes = 0)
opp_cols <- paste0("opp", 1:7)

results <- tournament %>%
  rowwise() %>%
  mutate(
    expected_score = {
      opps <- c_across(all_of(opp_cols))
      opps <- opps[opps != 0]  # remove byes
      r_player <- pre_ratings[as.character(id)]
      sum(sapply(opps, function(opp_id) {
        r_opp <- pre_ratings[as.character(opp_id)]
        elo_expected(r_player, r_opp)
      }))
    },
    pre_rating = pre_ratings[as.character(id)]
  ) %>%
  ungroup() %>%
  mutate(
    expected_score = round(expected_score, 2),
    difference     = round(actual_pts - expected_score, 2)
  ) %>%
  select(id, name, state, pre_rating, actual_pts, expected_score, difference) %>%
  arrange(desc(difference))
```

---

## Top 5 Overperformers

Players whose actual score most **exceeded** their Elo-expected score.

```{r overperformers}
top5_over <- results %>%
  slice_max(difference, n = 5)

top5_over %>%
  select(Name = name, State = state, `Pre-Rating` = pre_rating,
         `Actual Score` = actual_pts, `Expected Score` = expected_score,
         `Difference` = difference) %>%
  kable(align = c("l","c","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(6, bold = TRUE, color = "white", background = "#27ae60")
```

---

## Top 5 Underperformers

Players whose actual score most **fell short of** their Elo-expected score.

```{r underperformers}
top5_under <- results %>%
  slice_min(difference, n = 5)

top5_under %>%
  select(Name = name, State = state, `Pre-Rating` = pre_rating,
         `Actual Score` = actual_pts, `Expected Score` = expected_score,
         `Difference` = difference) %>%
  kable(align = c("l","c","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(6, bold = TRUE, color = "white", background = "#c0392b")
```

---

## Full Rankings

All 66 players sorted by performance difference (overperformers first).

```{r full-table}
results %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Name = name, State = state, `Pre-Rating` = pre_rating,
         `Actual` = actual_pts, `Expected` = expected_score,
         `Diff` = difference) %>%
  kable(align = c("r","l","c","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 13) %>%
  row_spec(which(results$difference >= 1.0),  background = "#1a3a1a", color = "white") %>%
  row_spec(which(results$difference <= -1.5), background = "#3a1a1a", color = "white")
```

---

## Bar Chart: Performance vs. Expectation

```{r bar-chart, fig.width=10, fig.height=12}
library(ggplot2)

results_plot <- results %>%
  mutate(name = factor(name, levels = rev(name)),
         fill_color = case_when(
           difference >= 1.0  ~ "Strong Over",
           difference > 0     ~ "Slight Over",
           difference > -1.0  ~ "Slight Under",
           TRUE               ~ "Strong Under"
         ))

ggplot(results_plot, aes(x = difference, y = name, fill = fill_color)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, color = "white", linewidth = 0.4) +
  scale_fill_manual(values = c(
    "Strong Over"  = "#27ae60",
    "Slight Over"  = "#82c99a",
    "Slight Under" = "#e08080",
    "Strong Under" = "#c0392b"
  )) +
  labs(
    title    = "Chess Tournament: Actual vs. Expected Score (Elo)",
    subtitle = "Difference = Actual Score − Expected Score",
    x        = "Difference (points)",
    y        = NULL,
    fill     = NULL,
    caption  = "Formula: E = 1 / (1 + 10^((R_opp − R_player)/400))\nSource: Elo (1978); Wikipedia — Elo rating system"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.background  = element_rect(fill = "#1a1a2e", color = NA),
    panel.background = element_rect(fill = "#1a1a2e", color = NA),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#333355"),
    panel.grid.minor   = element_blank(),
    text               = element_text(color = "#e8e0d0"),
    axis.text          = element_text(color = "#aaaaaa", size = 8),
    plot.title         = element_text(size = 13, face = "bold", color = "#c8a96e"),
    plot.subtitle      = element_text(size = 10, color = "#888888"),
    plot.caption       = element_text(size = 8,  color = "#555555"),
    legend.position    = "top",
    legend.text        = element_text(size = 9)
  )
```

---


