---
title: 'Project 1: Chess Tournament Data Transformation'
author: "Nana Kwasi Danquah"
date: '`r Sys.Date()`'
output: html_document
---

This project involves transforming a semi-structured chess tournament text file into a clean, tabular CSV format using R Markdown.

## My Approach

To fulfill this assignment, I have designed a step-by-step workflow in R to parse the semi-structured tournamentinfo.txt file.

### 1. Data Cleaning

I first read the raw text and removed the decorative dashed lines that act as separators. I also strip out the initial header rows to isolate the player data.

### 2. Structural Mapping

I observe that each player's data is split across two rows. I use indexing to separate these into two distinct vectors: one for primary data (Name and Points) and one for secondary data (State and Rating).

### 3. Regex Extraction

I use Regular Expressions to target specific fields:

**Name:** Characters between the first and second pipe symbols.

**State:** The two uppercase letters starting the second row.

**Total Points:** The numeric value in the "Total" column.

**Pre-Rating:** The digits immediately following "R:".

### 4. Averaging Opponents

I extract the numeric IDs of every opponent a player faced. I then use these IDs to look up their corresponding pre-ratings and calculate the mean for each player.

```{r setup, include=FALSE}
# I use tidyverse for string manipulation and data framing
library(tidyverse)
library(stringr)
```

## 1. Load and Clean Data

I read the file and remove the non-data elements like dashed lines and headers.
```{r}
# Read the provided text file
raw_txt <- readLines("tournamentinfo.txt")

# Remove dashed separator lines
clean_txt <- raw_txt[!str_detect(raw_txt, "^-+$")]

# Remove the header rows
clean_txt <- clean_txt[-(1:2)]
```

## 2. Separate Player Rows

Since each player has two rows of data, I separate them into two vectors.
```{r}
# Line 1 contains Name and Points
line1 <- clean_txt[seq(1, length(clean_txt), 2)]

# Line 2 contains State and Pre-Rating
line2 <- clean_txt[seq(2, length(clean_txt), 2)]
```

## 3. Extract Fields Using Regex

I extract the five required data points for my final CSV.
```{r}
# Extract Name, State, and Points
player_name  <- str_trim(str_extract(line1, "(?<=\\d\\s\\|\\s)[^|]+"))
player_state <- str_extract(line2, "[A-Z]{2}")
total_pts    <- as.numeric(str_extract(line1, "\\d+\\.\\d"))

# Extract Pre-Rating (handling the 'R:' prefix and potential provisional 'P' ratings)
pre_rating   <- as.numeric(str_extract(str_extract(line2, "R:\\s*\\d+"), "\\d+"))

# Extract Opponent IDs for the average calculation
# I look for the ID numbers following the results W, L, or D
opp_ids <- str_extract_all(line1, "(W|L|D)\\s+(\\d+)") %>%
           map(~ as.numeric(str_extract(.x, "\\d+")))
```

## 4. Calculate Average Opponent Rating

I look up each opponent's rating and calculate the mean for every player.
```{r}
avg_opp_rating <- sapply(opp_ids, function(ids) {
  round(mean(pre_rating[ids], na.rm = TRUE))
})
```

## 5. Final Table and CSV Export

I combine my results and generate the required file
```{r}
chess_df <- data.frame(
  Name = player_name,
  State = player_state,
  Total_Points = total_pts,
  Pre_Rating = pre_rating,
  Avg_Opp_Rating = avg_opp_rating
)

# Export to CSV for submission
write.csv(chess_df, "chess_tournament_results.csv", row.names = FALSE)

# Display the first few rows for validation
knitr::kable(head(chess_df))
```

