---
title: "Assignment — Working with HTML and JSON"
author: "Nana Kwasi Danquah"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
---

## Approach

### Subject and Book Selection

The subject chosen is **Science Fiction**. These books have clean, consistent
metadata that works well in both an HTML table and a JSON file.

The three books are:

1. **Dune** by Frank Herbert, published in 1965 by Chilton Books
2. **The Long War** by Terry Pratchett and Stephen Baxter, published in 2013 by Harper Collins — this is the multi-author book
3. **Neuromancer** by William Gibson, published in 1984 by Ace Books

Each book is recorded with five attributes: title, authors, year, publisher,
and ISBN. The authors field stores one or more names as a single
comma-separated string so the table structure stays consistent across all
three rows.

---

### Creating the Data Files

**books.html**

The HTML file is written by hand. It has a simple table with a header row
and three data rows, one per book. The five column headers match the five
attributes listed above. A small amount of CSS is added to make the table
easier to read, but there is no JavaScript and no external libraries — just
plain HTML. The file is hosted on GitHub so R can fetch it by URL.

**books.json**

The JSON file is also written by hand. It is an array of three objects where
each object has the same five keys as the HTML table columns. The year field
is stored as a number rather than a string because JSON supports that
natively. Everything else is stored as a string. The file lives in the same
GitHub repository as the HTML file.

---

## The Data Files

### books.html

Below is the raw HTML written by hand. The table has five columns and three
data rows. The structure is intentionally simple so the focus stays on the
data rather than the formatting.

```html
<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <title>Science Fiction Books</title>
</head>
<body>
  <table border="1" cellpadding="8" cellspacing="0">
    <thead>
      <tr>
        <th>Title</th>
        <th>Authors</th>
        <th>Year</th>
        <th>Publisher</th>
        <th>ISBN</th>
      </tr>
    </thead>
    <tbody>
      <tr>
        <td>Dune</td>
        <td>Frank Herbert</td>
        <td>1965</td>
        <td>Chilton Books</td>
        <td>978-0-441-17271-9</td>
      </tr>
      <tr>
        <td>The Long War</td>
        <td>Terry Pratchett, Stephen Baxter</td>
        <td>2013</td>
        <td>Harper Collins</td>
        <td>978-0-06-206726-8</td>
      </tr>
      <tr>
        <td>Neuromancer</td>
        <td>William Gibson</td>
        <td>1984</td>
        <td>Ace Books</td>
        <td>978-0-441-56956-4</td>
      </tr>
    </tbody>
  </table>
</body>
</html>
```

### books.json

Below is the raw JSON written by hand. Year is stored as a number. Every
other field is a string. The key names are lowercase to match what R expects
after cleaning the HTML column headers.

```json
[
  {
    "title": "Dune",
    "authors": "Frank Herbert",
    "year": 1965,
    "publisher": "Chilton Books",
    "isbn": "978-0-441-17271-9"
  },
  {
    "title": "The Long War",
    "authors": "Terry Pratchett, Stephen Baxter",
    "year": 2013,
    "publisher": "Harper Collins",
    "isbn": "978-0-06-206726-8"
  },
  {
    "title": "Neuromancer",
    "authors": "William Gibson",
    "year": 1984,
    "publisher": "Ace Books",
    "isbn": "978-0-441-56956-4"
  }
]
```

---

## Loading and Comparing the Data

### Step 1 — Load the HTML File

The `rvest` package is used to read the HTML file. `read_html` fetches the
page from GitHub, `html_element` finds the table, and `html_table` converts
it into a data frame. Column names are lowercased to make them consistent.
The year column comes in as character because `html_table` reads every cell
as text, so it is converted to integer right away.

```{r}
#| label: load-html
#| message: false

library(rvest)
library(dplyr)

html_url <- "https://raw.githubusercontent.com/nanadanquah05-jpg/DATA-607/refs/heads/main/books.html"

df_html <- read_html(html_url) |>
  html_element("table") |>
  html_table()

colnames(df_html) <- tolower(colnames(df_html))

df_html$year <- as.integer(df_html$year)

df_html
```

### Step 2 — Load the JSON File

The `jsonlite` package reads the JSON file. `fromJSON` parses the array
directly into a data frame. The year field is already an integer because
JSON stores it as a number, so no conversion is needed there.

```{r}
#| label: load-json
#| message: false

library(jsonlite)

json_url <- "https://raw.githubusercontent.com/nanadanquah05-jpg/DATA-607/refs/heads/main/books.json"

df_json <- fromJSON(json_url)

colnames(df_json) <- tolower(colnames(df_json))

df_json
```

### Step 3 — Compare the Two Data Frames

The comparison is done in two ways. First, a column-level summary shows the
data type from each source and whether the values match row for row. Second,
`identical` is called on both data frames and the result is printed as a
plain message.

```{r}
#| label: compare

# Align column order to match the HTML data frame
df_json <- df_json[, colnames(df_html)]

# Column-level summary
comparison <- data.frame(
  column       = colnames(df_html),
  html_type    = sapply(df_html, class),
  json_type    = sapply(df_json, class),
  values_match = mapply(function(a, b) all(a == b), df_html, df_json)
)

comparison
```

```{r}
#| label: identical-check

result <- identical(df_html, df_json)

if (result) {
  cat("The two data frames are identical.\n")
} else {
  cat("The two data frames are not identical.\n")
}
```

---

## Conclusion

Both files carry exactly the same information and the comparison confirms
they are identical after the year column in the HTML data frame is converted
to integer.

The main practical difference between the two formats is how they handle
data types. JSON can express numbers as numbers, so year arrives in R already
typed correctly. HTML has no type system — everything in a table cell is
text, so the type has to be assigned manually after loading.

In terms of structure, HTML is more verbose because the data is wrapped in
tags that describe how it should be displayed. JSON is more compact and
closer to how data is actually stored and exchanged between systems. For
loading data into R, JSON is simpler because `fromJSON` handles almost
everything automatically. With HTML, a bit more cleanup is needed after the
table is extracted.

Both formats are worth knowing. HTML tables appear all over the web and
being able to pull them into R with `rvest` is a practical skill. JSON is
the standard format for most web APIs, so `jsonlite` comes up constantly in
real data work.
